TRIK Studio: Technical Introduction

### 2017-02-09 2017-trikStudioEnglish/trikStudioEnglish.tex
translated interpreters

\begin{abstract}
This paper presents TRIK Studio --- an environment for visual (and textual) programming of robotic kits, which is used in educational organizations across Russia and Europe. First part of the article provides overview of the system --- its purpose, features, differences from similar programming environments, general difficulties of robot programming and solutions proposed by TRIK Studio. Second part presents implementation details of TRIK Studio and its most interesting components. This article combines five fields of study: robotics, domain-specific visual modelling, education, formal methods and methods of program analysis. Main contribution of this article is detailed technical description of TRIK Studio as a complex and successful open-source cross-platform robot programming environment written in C++/Qt, and first part of the article can also be interesting for teachers as it provides an overview of existing robot programming tools and related problems.
\end{abstract}

\section{Introduction}
\label{chapter:intro}
Current state of school education in computer science turned out to be quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowdays. It is less known that Papert in his experiments also used a mechanical robot turtle, that was controlled from a computer ~\cite{papert1980mindstorms}, and it made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course \cite{черёмухин2014внедрение,лучин2016внедрение}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3 \footnote{LEGO Mindstorms homepage, URL: http://www.lego.com/en-us/mindstorms (accessed: 07.02.2017)}, 
TRIK\footnote{TRIK robotics platform homepage, URL: http://www.trikset.com/ (accessed: 07.02.2017)} etc.

@@

For the last decade educational robotics has been and still is a very promising scientific field, so in 2010s almost every major university in the world invested in development of own projects in robotics. For example, in 2016 Harvard University presented Root platform~\cite{root}, Carnegie Mellon University develops and promotes its Arts\&Bots project~\cite{cross2013visual}, MIT contributed delivering Scratch environment~\cite{resnick2009scratch} and so on. Detailed overview in Russian could be found in~\cite{mordvinov2016NONPUBLISHED}, and the author concludes there that despite the variety of tools in this field there is no single one that can satisfy the needs of all educational organizations. Vast majority of such tools implement only the basic functionality (visual editor for diagrams and an ability to run programs on a robot autonomously or in a controlled manner from the computer), but they lack more advanced tools for teaching programming. Examples of such tools could be generation of readable texual code from created diagrams (that will help students to migrate from diagram-based to code-based programs), tools for debugging the program using virtual robot simulator or embedded tools for checking correctness of created programs (which could help teachers checking students' tasks). Some tools do have some of these advanced features, but often only some of them, and most of such programming environments are proprietary (which puts their users into a vendor lock-in situation) and pretty expensive for lots of schools.

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to understand, so it suits the purpose of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: the program controlling the robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where the program consists of a number of <<black boxes>>, connected with data channels. Each such <<box>> (we will call it \textit{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call data transfering through channels \textit{tokens>}). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sendors data to its output. Many researchers  note usability of visual data-flow langiages comparing to texual ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

The idea of using data-flow programming languages in robotics is not a novel one, they are implemented in almost every toolset for programming industrial and laboratory automation system. Among them are LabVIEW by National Instruments~\cite{kodosky1991visual}, Simulink~\cite{dabney2004mastering}, and Microsoft Robotics Developer Studio~\cite{jackson2007microsoft}. All of them solve the automation task very well and have very powerful tools inside (for example, Microsoft Robotics Developer Studio is run on MySpace social network servers~\cite{scherotter2009ccr}), but are very complex and hard to learn, so even if they are used in education, only in universities~\cite{stefanovic2011labview,yi2005labview}. School experiments with LabVIEW were popular in the late 1990s~\cite{cyr1997low,portsmore1999robolab}, which resulted in adaptations of this environment (like Robolab), where data-flow model was replaced with control-flow model. So, a gap between simplified languages suitable for education and more elegant, but complex industrial languages exists. Some reseachers are trying to adapt the data-flow model for educational robotics, for example a research group from New Zealand introduced the RuRu~\cite{diprose2011ruru} language. Nevertheless to our best knowledge there are still no production-ready data-flow based programming environments for educational purpose today.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It 	emerged as a further evolution of QReal:Robots~\cite{terekhov2013sreda} project, developed at Software Engineering chair of Saint-Petersburg State University. The official release supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (the more simple control-flow and more complex data-flow language) or one of a number of texual ones. For Lego NXT the programmer can choose from NXT OSEK C and Russian version of C language (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual byte code, is supported.

@@

The the first mode the program is being interpreted on a two-dimentional robot model (see Section~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot out of walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

@@

Code generation mode enables users to move from visual to textual programs. Gerenated code is displayed in an embedded QScintilla-based~\footnote{https://riverbankcomputing.com/software/qscintilla/intro (accessed: 07.02.2017)} text editor, that provides a set of common features of a code editor: syntax highlighting, code audocompletion, undo/redo, brackets highlighting etc. TRIK Studio installation contains everything neede for buiding and uploading programs into a robot (a number of cross-compilers, WinSCP, Putty, etc.), and it allows to make compilation process and communication with robots transparent to TRIK Studio users.

@@

Aforementioned features distinguish TRIK Studio from all other such tools. In fact among all mentioned in Section~\ref{chapter:intro} tools only 12Blocks have a comparable feature set, but it generates complex and unreadable code (for Lego NXT), does not have tools for checking tasks, is not free to use and does not have Russian localization in it. Among the TRIK Studio drawbacks we should mention weak methodological support, limited only by embedded reference guide in Russian, a set of examaples and the remote video course. Less important drawbacks will be mentioned in the following sections.

The rest of the paper is structured as follows. Sections~\ref{chapter:controlFlowLanguage} and~\ref{chapter:dataFlowLanguage} briefly present TRIK Studio visual languages and their interpreters. Section~\ref{chapter:commonArchitecture} overviews the system architecture. Further sections describe separate TRIK Studio subsystems. Section~\ref{chapter:communications} presents robots communication infrastructure. Section~\ref{chapter:interpreters} describes visual language interpreters. The most interesting implementation details of code generators from control-flow language are provided in Section~\ref{chapter:generators}. Section~\ref{chapter:parser} presents textual languages parsing subsystem. Section~\ref{chapter:2dModel} presents implementation details of 2D simulation model sybsystem. Section~\ref{chapter:constraintsChecker} describes task checking language and Section~\ref{chapter:conclusion} concludes the paper.

@@

The simplified control-flow based lanaguage (see Fig.~\ref{image:TS_CF_Example}) is the most often used one in TRIK Studio. It it a graph language, i.e. the program consists of nodes (\textit{blocks}) and edges, organizing the nodes into a control flow. To create a program users drag-and-drop necesary blocks onto the editor's scene, sets their properties and connects them with arrows. While being run each block executes a sequence of basic commands and passed control to outgoing edges (to all or some of them depending on the block semantics).

@@

\begin{itemize}
    \item The first group is for blocks supporting basic algotithmic expressions, like the start and the end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for execution parallelization and for working with concurrent tasks (e.g. for merging and interruprting tasks), subprogram call and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, they are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that <<freeze>> the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator gamepad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text, images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

@@

Properties of each block can be set both within the diagram editor's scene and using special property editor window. All block properties (where applicable) support computable expressions written in TRIK Studio embedded textual language -- a statically typed Lua\footnote{http://www.lua.ru/ (accessed: 07.02.2017)} dialect. The parsing module for this parguage is written using a parser combinator library for C++11, also created within TRIK Studio project. Type inference of the resulting abstract syntax tree is performed using slightly simplified Hindley-Milner algorithm~\cite{damas1982principal}.

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. An editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}.  Language metamodel was defined using QReal's metaeditor, and a plugin implementing the editor was automatically generated using QReal's tools. Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE <<inherits>> all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~\cite{osechkina2010gestures,osechkina2012multistroke}, copy-pasting and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent about three man-days on it. We believe that it acknoledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

Users that mastered the control-flow language can move on to more complex, but more convenient data-flow visual language. Unlike control-flow programs where control is passed according to edges between nodes, in data-flow programs blocks are executed simultaniously and communicate with each other by sending data tokens via channels. For instance, a data-flow program can have several entry points (most often they are robot sensors that send data each other for processing in a chain), while a control-flow program must always have a single entry point. This language also has block for basic algotithmic expressions, blocks for interaction with all robot devices, drawing blocks, etc. Almost always data-flow programs turn out to be more concise than their control-flow analogues. For example, Fig.~\ref{image:alongTheBox_CF_DF} shows proportional controller implemented in control-flow language is compared with proportional-derivative controller implemented in data-flow language.

@@

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's <<colony>> architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not given here. Another ideas on this matter could be found, for exmaple, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Fig.~\ref{image:commonTSArch} presents high-level architecture of TRIK Studio. The system has a number of layers, each one implementing specific set of functionality and having strictly defined API. The lowest layer is libraries implementing interaction with real robots and and virtual robot models. They are used to implement a hierarchy of robot devices (sensors, motors, displays, speakers, gamepads, control buttons, etc.), which in their turn are used to specify high-level robot models for different robotic kits. Such robot model specifications are grouped into modules that are plugged into TRIK Studio's core, where they are used by other subsystems (e.g. intrepreters and code generators that are structured as plug-in modules too).

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Futhermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not <<know>> about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extention modules supporting new languages, robotic kits and tools.

@@

\begin{itemize}
    \item changing current transmission medium (currently Bluetooth, Wi-Fi and USB are supported),
    \item setting device address in terms of a chosen medium (e.g. COM port number in case of Bluetooth, IP address or host name in case of Wi-Fi),
    \item connecting to and disconnrect from a remote device,
    \item transfering byte arrays to a remove device,
    \item handling recieving data from a remove device,
    \item handling connection status change (i.e. connection setup or termination events),
    \item handling errors.
\end{itemize}

USB communication is implemented using libusb\footnote{http://libusb.org/ (accessed: 07.02.2017)} library, Bluetooth communication is based on QextSerialPort\footnote{https://github.com/qextserialport/ (accessed: 07.02.2017)} library, Wi-Fi communication is implemented over TCP and UDP protocols using QtNetwork library, complex communication protocols (like configuring TRIK robots before running programs) are built upon Qt State Machine Framework. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implemetations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

Интерпретатор преобразует визуальную диаграмму в последовательность команд на целевом устройстве (рис.~\ref{image:interpretersTSArch}). При этом иерархия устройств в системе построена таким образом, что не делается различий, реальное ли это устройство или симулируемое (рис.~\ref{image:devicesTSArch}). Каждое устройство представляет собой класс C++, содержащий код взаимодействия с целевым механизмом. Реализации устройств какого-либо конструктора находятся в соответствующих библиотеках подключаемых модулей. Для общения по физическим каналам устройства используют подсистему коммуникаций (гл.~\ref{chapter:communications}), для передачи команд двумерному и трехмерному симуляторам --- API этих симуляторов.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\columnwidth]{TS_Interpreter_Architecture.png}
    \caption{Общий принцип работы систем интерпретации}
    \label{image:interpretersTSArch}
\end{figure}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.7\textwidth]{TS_Devices_Architecture.png}
    \caption{Часть иерархии устройств в системе}
    \label{image:devicesTSArch}
\end{figure*}

Интерпретаторы, также как и вся система в целом, написан на языке C++ с использованием инструментария Qt. На вход подсистеме интерпретации поступает
описание созданной пользователем диаграммы в некотором внутреннем представлении. В среде реализовано два интерпретатора: интерпретатор языка для начинающих с передачей управления и интерпретатор потокового языка.

Интерпретатор первого языка находит на диаграмме начальный блок, и далее трассирует поток управления диаграммы в том порядке, как он задан стрелками. Текущий блок исполнения подсвечивается на визуальной диаграмме, чтобы пользователю был виден прогресс исполнения. Для каждого посещенного блока специальные фабрики интерпретатора создают объекты, которые реализуют его поведение. В общем случае, такой объект-реализация ищет среди готовых к работе устройств робота необходимые ему, вызывает соответствующие команды и передает управление следующим блокам по какой-либо исходящей ветке (которая определяется, опять-таки, самой реализацией блока). При этом не делается различий, является ли найденное устройство частью реального робота или симулируемым, таким образом, одна и та же подсистема интерпретации используется для исполнения диаграммы в двух режимах из трех (гл.~\ref{chapter:generalDescription}).

Если на каком-то шаге интерпретации повстречался блок распараллеливания, интерпретатор запускает новые потоки исполнения с соответствующих точек, создавая для каждого свой стек вызовов. Если встретился вызов подпрограммы, вычисляются ее параметры, укладываются на текущий стек вызовов, и далее процесс интерпретации повторяется рекурсивно. При достижении любого завершающего блока со стека снимается верхний фрейм, и исполнение продолжается с соответствующей точки на предыдущей диаграмме. Если при снятии очередного фрейма стек вызовов стал пустым, текущий поток исполнения считается завершенным. Таким образом, программа исполняется по мере того, как она <<открывается>> интерпретатору.

>>>>

\begin{abstract}
This paper presents TRIK Studio --- an environment for visual (and textual) programming of robotic kits, which is used in educational organizations across Russia and Europe. First part of the article provides overview of the system --- its purpose, features, differences from similar programming environments, general difficulties of robot programming and solutions proposed by TRIK Studio. Second part presents implementation details of TRIK Studio and its most interesting components. This article combines five fields of study: robotics, domain-specific visual modeling, education, formal methods and methods of program analysis. Main contribution of this article is detailed technical description of TRIK Studio as a complex and successful open-source cross-platform robot programming environment written in C++/Qt, and first part of the article can also be interesting for teachers as it provides an overview of existing robot programming tools and related problems.
\end{abstract}

\section{Introduction}
\label{chapter:intro}
Current state of school education in computer science turned out to be quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowadays. It is less known that Papert in his experiments also used a mechanical robot turtle, that was controlled from a computer ~\cite{papert1980mindstorms}, and it made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course \cite{черёмухин2014внедрение,лучин2016внедрение}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3 \footnote{LEGO Mindstorms homepage, URL: http://www.lego.com/en-us/mindstorms (accessed: 07.02.2017)}, 
TRIK\footnote{TRIK robotics platform homepage, URL: http://www.trikset.com/ (accessed: 07.02.2017)} etc.

@@

For the last decade educational robotics has been and still is a very promising scientific field, so in 2010s almost every major university in the world invested in development of own projects in robotics. For example, in 2016 Harvard University presented Root platform~\cite{root}, Carnegie Mellon University develops and promotes its Arts\&Bots project~\cite{cross2013visual}, MIT contributed delivering Scratch environment~\cite{resnick2009scratch} and so on. Detailed overview in Russian could be found in~\cite{mordvinov2016NONPUBLISHED}, and the author concludes there that despite the variety of tools in this field there is no single one that can satisfy the needs of all educational organizations. Vast majority of such tools implement only the basic functionality (visual editor for diagrams and an ability to run programs on a robot autonomously or in a controlled manner from the computer), but they lack more advanced tools for teaching programming. Examples of such tools could be generation of readable textual code from created diagrams (that will help students to migrate from diagram-based to code-based programs), tools for debugging the program using virtual robot simulator or embedded tools for checking correctness of created programs (which could help teachers checking students' tasks). Some tools do have some of these advanced features, but often only some of them, and most of such programming environments are proprietary (which puts their users into a vendor lock-in situation) and pretty expensive for lots of schools.

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to understand, so it suits the purpose of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: the program controlling the robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where the program consists of a number of <<black boxes>>, connected with data channels. Each such <<box>> (we will call it \textit{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call data transferring through channels \textit{tokens>}). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sensors data to its output. Many researchers  note usability of visual data-flow languages comparing to textual ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

The idea of using data-flow programming languages in robotics is not a novel one, they are implemented in almost every toolset for programming industrial and laboratory automation system. Among them are LabVIEW by National Instruments~\cite{kodosky1991visual}, Simulink~\cite{dabney2004mastering}, and Microsoft Robotics Developer Studio~\cite{jackson2007microsoft}. All of them solve the automation task very well and have very powerful tools inside (for example, Microsoft Robotics Developer Studio is run on MySpace social network servers~\cite{scherotter2009ccr}), but are very complex and hard to learn, so even if they are used in education, only in universities~\cite{stefanovic2011labview,yi2005labview}. School experiments with LabVIEW were popular in the late 1990s~\cite{cyr1997low,portsmore1999robolab}, which resulted in adaptations of this environment (like Robolab), where data-flow model was replaced with control-flow model. So, a gap between simplified languages suitable for education and more elegant, but complex industrial languages exists. Some researchers are trying to adapt the data-flow model for educational robotics, for example a research group from New Zealand introduced the RuRu~\cite{diprose2011ruru} language. Nevertheless to our best knowledge there are still no production-ready data-flow based programming environments for educational purpose today.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It 	emerged as a further evolution of QReal:Robots~\cite{terekhov2013sreda} project, developed at Software Engineering chair of Saint-Petersburg State University. The official release supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (the more simple control-flow and more complex data-flow language) or one of a number of textual ones. For Lego NXT the programmer can choose from NXT OSEK C and Russian version of C language (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual byte code, is supported.

@@

The the first mode the program is being interpreted on a two-dimensional robot model (see Section~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot out of walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

@@

Code generation mode enables users to move from visual to textual programs. Generated code is displayed in an embedded QScintilla-based~\footnote{https://riverbankcomputing.com/software/qscintilla/intro (accessed: 07.02.2017)} text editor, that provides a set of common features of a code editor: syntax highlighting, code autocompletion, undo/redo, brackets highlighting etc. TRIK Studio installation contains everything needed for building and uploading programs into a robot (a number of cross-compilers, WinSCP, Putty, etc.), and it allows to make compilation process and communication with robots transparent to TRIK Studio users.

@@

Aforementioned features distinguish TRIK Studio from all other such tools. In fact among all mentioned in Section~\ref{chapter:intro} tools only 12Blocks have a comparable feature set, but it generates complex and unreadable code (for Lego NXT), does not have tools for checking tasks, is not free to use and does not have Russian localization in it. Among the TRIK Studio drawbacks we should mention weak methodological support, limited only by embedded reference guide in Russian, a set of examples and the remote video course. Less important drawbacks will be mentioned in the following sections.

The rest of the paper is structured as follows. Sections~\ref{chapter:controlFlowLanguage} and~\ref{chapter:dataFlowLanguage} briefly present TRIK Studio visual languages and their interpreters. Section~\ref{chapter:commonArchitecture} overviews the system architecture. Further sections describe separate TRIK Studio subsystems. Section~\ref{chapter:communications} presents robots communication infrastructure. Section~\ref{chapter:interpreters} describes visual language interpreters. The most interesting implementation details of code generators from control-flow language are provided in Section~\ref{chapter:generators}. Section~\ref{chapter:parser} presents textual languages parsing subsystem. Section~\ref{chapter:2dModel} presents implementation details of 2D simulation model subsystem. Section~\ref{chapter:constraintsChecker} describes task checking language and Section~\ref{chapter:conclusion} concludes the paper.

@@

The simplified control-flow based language (see Fig.~\ref{image:TS_CF_Example}) is the most often used one in TRIK Studio. It it a graph language, i.e. the program consists of nodes (\textit{blocks}) and edges, organizing the nodes into a control flow. To create a program users drag-and-drop necessary blocks onto the editor's scene, sets their properties and connects them with arrows. While being run each block executes a sequence of basic commands and passed control to outgoing edges (to all or some of them depending on the block semantics).

@@

\begin{itemize}
    \item The first group is for blocks supporting basic algorithmic expressions, like the start and the end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for parallel execution and for working with concurrent tasks (e.g. for merging and interrupting tasks), subprogram call and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, they are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that <<freeze>> the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator game pad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text, images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

@@

Properties of each block can be set both within the diagram editor's scene and using special property editor window. All block properties (where applicable) support computable expressions written in TRIK Studio embedded textual language -- a statically typed Lua\footnote{http://www.lua.ru/ (accessed: 07.02.2017)} dialect. The parsing module for this language is written using a parser combinator library for C++11, also created within TRIK Studio project. Type inference of the resulting abstract syntax tree is performed using slightly simplified Hindley-Milner algorithm~\cite{damas1982principal}.

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. An editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}.  Language metamodel was defined using QReal's metaeditor, and a plug-in implementing the editor was automatically generated using QReal's tools. Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE <<inherits>> all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~\cite{osechkina2010gestures,osechkina2012multistroke}, copy-pasting and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent about three man-days on it. We believe that it acknowledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

Users that mastered the control-flow language can move on to more complex, but more convenient data-flow visual language. Unlike control-flow programs where control is passed according to edges between nodes, in data-flow programs blocks are executed simultaneously and communicate with each other by sending data tokens via channels. For instance, a data-flow program can have several entry points (most often they are robot sensors that send data each other for processing in a chain), while a control-flow program must always have a single entry point. This language also has block for basic algorithmic expressions, blocks for interaction with all robot devices, drawing blocks, etc. Almost always data-flow programs turn out to be more concise than their control-flow analogues. For example, Fig.~\ref{image:alongTheBox_CF_DF} shows proportional controller implemented in control-flow language is compared with proportional-derivative controller implemented in data-flow language.

@@

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's <<colony>> architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not given here. Another ideas on this matter could be found, for example, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Fig.~\ref{image:commonTSArch} presents high-level architecture of TRIK Studio. The system has a number of layers, each one implementing specific set of functionality and having strictly defined API. The lowest layer is libraries implementing interaction with real robots and and virtual robot models. They are used to implement a hierarchy of robot devices (sensors, motors, displays, speakers, game pads, control buttons, etc.), which in their turn are used to specify high-level robot models for different robotic kits. Such robot model specifications are grouped into modules that are plugged into TRIK Studio's core, where they are used by other subsystems (e.g. interpreters and code generators that are structured as plug-in modules too).

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Furthermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not <<know>> about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extension modules supporting new languages, robotic kits and tools.

@@

\begin{itemize}
    \item changing current transmission medium (currently Bluetooth, Wi-Fi and USB are supported),
    \item setting device address in terms of a chosen medium (e.g. COM port number in case of Bluetooth, IP address or host name in case of Wi-Fi),
    \item connecting to and disconnect from a remote device,
    \item transferring byte arrays to a remove device,
    \item handling receiving data from a remove device,
    \item handling connection status change (i.e. connection setup or termination events),
    \item handling errors.
\end{itemize}

USB communication is implemented using libusb\footnote{http://libusb.org/ (accessed: 07.02.2017)} library, Bluetooth communication is based on QextSerialPort\footnote{https://github.com/qextserialport/ (accessed: 07.02.2017)} library, Wi-Fi communication is implemented over TCP and UDP protocols using QtNetwork library, complex communication protocols (like configuring TRIK robots before running programs) are built upon Qt State Machine Framework. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implementations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

An interpreter is translating visual diagrams into a sequence of command for a target device (see Fig.~\ref{image:interpretersTSArch}). The hierarchy of devices is build in a way that it does not matter to an interpreter if it works with a real device of a virtual one (see Fig.~\ref{image:devicesTSArch}). Each device is represented by a C++ class containing code for interaction with this device. These device classes are grouped according to robotic kits they belong to and put in plug-in modules. Device classes use communication subsystem for data interchange over physical channels (see Section~\ref{chapter:communications}) and 2D and 3D simulators API when working with virtual robot models.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\columnwidth]{TS_Interpreter_Architecture.png}
    \caption{Diagram interpretation workflow}
    \label{image:interpretersTSArch}
\end{figure}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.7\textwidth]{TS_Devices_Architecture.png}
    \caption{TRIK Studio's device classes hierarchy}
    \label{image:devicesTSArch}
\end{figure*}

All interpreters, like all the rest of the environment, is written in C++/Qt. As input data they receive created diagrams in some external representation format. There are two interpreters implemented in TRIK Studio now: one for control-flow programs and one for data-flow programs.

The interpreter of control-flow programs firstly starts with finding initial block of the program and then iterates according to the control flow defined by the diagram arrows. Currently executed block is highlighted in the editor to show users the execution process. For each block visited special factories create objects that implement the behavior of this block. In general such an object searches for a ready-to-work robot device that it needs, runs appropriate commands and passes execution to one of the blocks that the current block has outgoint links to (depending of the block's semantics). In this process no difference is made if the device is a real one or a virtual one, so the same interpretation code works for execution of the program in two out of three modes (see Section~\ref{chapter:generalDescription}).

If on any step of the interpretation process a fork block is met, the interpreter launches new threads of execution, creating a new call stack for each of them. If a subprogram call block is met, its arguments are computed and put into the current call stack, and the interpretation process is repeated recursively. When processing any type of final block the interpreter removes the frame from the call stack and the execution is being continued from the proper point of the previous diagram. If we get an empty call stack after removing the frame, the current execution thread is considered finished. In this way the interpretation process is going on while the interpreter can find new blocks to process.

An interesting feature of this implementation is that users can change the diagram during its interpretation and the changes will be applied <<on the fly>>. It is a convenient behavior, for example, for selecting parameters of a PID controller in a program. Meanwhile, the execution of unchanged diagram parts is optimized: block implementation objects are cached, the same is true for abstract syntax trees and type inference information for textual programs. Thus, if the program remains unchanged while being interpreted, subsequent control flow to the same parts of the diagram will employ already created objects. Diagrams validation is also performed <<on the fly>>. If some incorrect fragment is found, an error message is shown. It could be seen as a drawback since users get their error messages only when such errors are reached during interpretation, but this behavior is typical for all widely used interpreted textual languages.

The data-flow interpreter works in two steps. First, the interpreted diagram is being validated, implementation objects for diagram blocks are created (they are used to translate the diagram into a sequence of robot commands similar to as it was discussed for the control-flow programs interpreter). Implementation objects are connected with each other via Qt signals and slots in the same order as corresponding blocks are connected on a diagram (<<Publisher-subscriber>> design pattern is used here). On the second step the interpreter executes each block that does not have any incoming data flows, and then each such block in its turn activates blocks that it has outgoing data flows to when the data tokens are ready. Blocks execution is performed in a pseudo-parallel way over a centralized message queue (see Fig.~\ref{image:dsPseudoparallelism}). This implementation detail distinguishes this language from all other industrial solutions (e.g. in Microsoft Robotics Developer Studio diagrams are deployed as a set of independent web services~\cite{jackson2007microsoft). We choose this approach because this programming environment targets low-performance controllers of educational robots, that don't embrace true parallel execution of multiple programs at once. Nevertheless, the language has a process fork block that is useful while implementing complex control models mentioned above. This block should be used as a low-level tool for execution control.

\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{DF_Interpretation_Process.png}
    \caption{Data-flow programs execution workflow}
    \label{image:dsPseudoparallelism}
\end{figure}
